---
title: "sts scoring"
author: "Daniel Zweben"
date: "2025-06-02"
output: html_document
---

```{r}

df <- read.csv("/Users/dannyzweben/Desktop/CABLAB_Files/Psychonomics_2025/Megan/sas-data/sts.csv")


```

####################### scoring 4 months of data screen time data - drop a score if its in the bottom 10th percentile for a month across all 3 (indicates they filled it in wrong and used today's data early in the morning or this weeks data on sunday....incorrect)

```{r}
library(dplyr)

# Step 1: Compute 10th percentile cutoffs
minutes_cutoff <- quantile(unlist(df[, c("total_minutes_1", "total_minutes_2", "total_minutes_3", "total_minutes_4")]), 0.10, na.rm = TRUE)
pickup_cutoff  <- quantile(unlist(df[, c("pickup1", "pickup2", "pickup3", "pickup4")]), 0.10, na.rm = TRUE)
notif_cutoff   <- quantile(unlist(df[, c("notif1", "notif2", "notif3", "notif4")]), 0.10, na.rm = TRUE)

# Step 2: Clean data
library(dplyr)


# Step 2: Conditional cleaning — drop only if all 3 at a timepoint are below threshold
df_clean <- df %>%
  mutate(
    clean_total_minutes_1 = ifelse(total_minutes_1 < minutes_cutoff & pickup1 < pickup_cutoff & notif1 < notif_cutoff, NA, total_minutes_1),
    clean_pickup1         = ifelse(pickup1 < pickup_cutoff & total_minutes_1 < minutes_cutoff & notif1 < notif_cutoff, NA, pickup1),
    clean_notif1          = ifelse(notif1 < notif_cutoff & total_minutes_1 < minutes_cutoff & pickup1 < pickup_cutoff, NA, notif1),

    clean_total_minutes_2 = ifelse(total_minutes_2 < minutes_cutoff & pickup2 < pickup_cutoff & notif2 < notif_cutoff, NA, total_minutes_2),
    clean_pickup2         = ifelse(pickup2 < pickup_cutoff & total_minutes_2 < minutes_cutoff & notif2 < notif_cutoff, NA, pickup2),
    clean_notif2          = ifelse(notif2 < notif_cutoff & total_minutes_2 < minutes_cutoff & pickup2 < pickup_cutoff, NA, notif2),

    clean_total_minutes_3 = ifelse(total_minutes_3 < minutes_cutoff & pickup3 < pickup_cutoff & notif3 < notif_cutoff, NA, total_minutes_3),
    clean_pickup3         = ifelse(pickup3 < pickup_cutoff & total_minutes_3 < minutes_cutoff & notif3 < notif_cutoff, NA, pickup3),
    clean_notif3          = ifelse(notif3 < notif_cutoff & total_minutes_3 < minutes_cutoff & pickup3 < pickup_cutoff, NA, notif3),

    clean_total_minutes_4 = ifelse(total_minutes_4 < minutes_cutoff & pickup4 < pickup_cutoff & notif4 < notif_cutoff, NA, total_minutes_4),
    clean_pickup4         = ifelse(pickup4 < pickup_cutoff & total_minutes_4 < minutes_cutoff & notif4 < notif_cutoff, NA, pickup4),
    clean_notif4          = ifelse(notif4 < notif_cutoff & total_minutes_4 < minutes_cutoff & pickup4 < pickup_cutoff, NA, notif4)
  )


df_result <- df_clean %>%
  rowwise() %>%
  mutate(
    avg_minutes_clean = mean(c_across(starts_with("clean_total_minutes_")), na.rm = TRUE),
    avg_pickups_clean = mean(c_across(starts_with("clean_pickup")), na.rm = TRUE),
    avg_notifs_clean  = mean(c_across(starts_with("clean_notif")), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  select(record_id, PID, avg_minutes_clean, avg_pickups_clean, avg_notifs_clean)


```

QULality control data wiped check: 
### checking if we ended up breaking up participant data - only happened a bit- looks good

```{r}
library(dplyr)

# This assumes you already have df (original) and df_clean (after applying cleaning steps)

df_qc <- df %>%
  select(PID, record_id,
         total_minutes_1:total_minutes_4,
         pickup1:pickup4,
         notif1:notif4) %>%
  mutate(
    # Had *any* raw values
    had_minutes_data = rowSums(!is.na(across(c(total_minutes_1, total_minutes_2, total_minutes_3, total_minutes_4)))) > 0,
    had_pickups_data = rowSums(!is.na(across(c(pickup1, pickup2, pickup3, pickup4)))) > 0,
    had_notifs_data  = rowSums(!is.na(across(c(notif1, notif2, notif3, notif4)))) > 0
  )

df_clean_flags <- df_clean %>%
  mutate(
    all_minutes_removed = rowSums(is.na(across(starts_with("clean_total_minutes_")))) == 4,
    all_pickups_removed = rowSums(is.na(across(starts_with("clean_pickup")))) == 4,
    all_notifs_removed  = rowSums(is.na(across(starts_with("clean_notif")))) == 4
  ) %>%
  select(PID, record_id, starts_with("all_"))

# Merge raw + clean flags
df_qc_flagged <- df_qc %>%
  left_join(df_clean_flags, by = c("PID", "record_id")) %>%
  mutate(
    minutes_removed_after_having_data = had_minutes_data & all_minutes_removed,
    pickups_removed_after_having_data = had_pickups_data & all_pickups_removed,
    notifs_removed_after_having_data  = had_notifs_data & all_notifs_removed
  ) %>%
  filter(minutes_removed_after_having_data |
         pickups_removed_after_having_data |
         notifs_removed_after_having_data) %>%
  select(PID, record_id,
         minutes_removed_after_having_data,
         pickups_removed_after_having_data,
         notifs_removed_after_having_data)

# View who's affected
print(df_qc_flagged)



df_clean %>%
  mutate(
    # Raw value counts
    n_minutes_raw  = rowSums(!is.na(df[, c("total_minutes_1", "total_minutes_2", "total_minutes_3", "total_minutes_4")])),
    n_pickups_raw  = rowSums(!is.na(df[, c("pickup1", "pickup2", "pickup3", "pickup4")])),
    n_notifs_raw   = rowSums(!is.na(df[, c("notif1", "notif2", "notif3", "notif4")])),

    # Cleaned value counts
    n_minutes_clean = rowSums(!is.na(across(starts_with("clean_total_minutes_")))),
    n_pickups_clean = rowSums(!is.na(across(starts_with("clean_pickup")))),
    n_notifs_clean  = rowSums(!is.na(across(starts_with("clean_notif"))))
  ) %>%
  summarise(
    minutes_under2 = sum(n_minutes_raw > 0 & n_minutes_clean < 2),
    pickups_under2 = sum(n_pickups_raw > 0 & n_pickups_clean < 2),
    notifs_under2  = sum(n_notifs_raw > 0 & n_notifs_clean < 2)
  )


```

##### how many avgs. do we have left?
##### looks like a lot - good to go ~50 of each score!!!!! ##########

```{r}
df_result %>%
  summarise(
    n_avg_minutes  = sum(!is.na(avg_minutes_clean)),
    n_avg_pickups  = sum(!is.na(avg_pickups_clean)),
    n_avg_notifs   = sum(!is.na(avg_notifs_clean))
  )

```


```{r}
write.csv(df_result, "/Users/dannyzweben/Desktop/CABLAB_Files/Psychonomics_2025/Megan/screen_time_data.csv", row.names = FALSE)

```


